---
title: "R Notebook"
output: html_notebook
---
```{r}
library(tidymodels)
library(tidyverse)
library(modelr)
library(rsample)
library(yardstick)
```

```{r}
# load the dataset
wd <- dirname(getwd())
load(paste(wd, "/data/biomarker-clean.Rdata", sep = ""))
```

```{r}
biomarker_clean_noAdos = biomarker_clean %>% select(-ados) 
```

## Split the data 

```{r}
# for reproducibility
set.seed(28292)

# partition data
partitions <- biomarker_clean %>%
  initial_split(prop = 0.8, strata = group)

# examine
asd_training = training(partitions)
asd_testing = testing(partitions)
```



## T-tests
$\mu^i_{ASD}$: mean serum level of protein $i$ in the ASD group 

$\mu^i_{TD}$: mean serum level of protein $i$ in the TD group

$\delta_i$: difference between $\mu^i_{ASD}$ - $\mu^i_{TD}$

Assumptions: 

$p$-values for a test is the probability of obtaining a sample at least as contrary to $H_{0i}$ as the sample in hand, assuming $H_{0i}$ is true.



Use **bonferroni correction**
  need explanation 

```{r}
asd_nested <- asd_training %>%
  select(-ados) %>%
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  nest(data = c(level, group))
```

```{r}
tt_fn <- function(.df){
  infer::t_test(.df, 
                formula = level ~ group,
                order = c('ASD', 'TD'),
                alternative = 'two-sided',
                var.equal = F)
}

n_tests <- nrow(asd_nested)

tt_out <- asd_nested %>%
  slice(1:n_tests) %>%
  mutate(ttest = map(data, tt_fn))
```

```{r}
test <-tt_out %>% 
  unnest(ttest) %>%
  select(protein, p_value) %>%
  arrange(p_value)
```

```{r}
t_result_2 <- test %>% 
  add_column(rank = 1:nrow(test)) %>%
  arrange(p_value) %>%
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)

t_test_proteins <- t_result_2 %>% 
  slice_min(p.adj, n = 15) %>%
  pull(protein)
#t_result_2 %>% filter(p_adj < 0.01) %>% select(protein) 
t_result_2
```
```{r}
t_test_proteins
```
```{r}
# save t_tests result

save(t_result_2, t_test_proteins, file = "t_tests_data.RData")
```



## Correlation Approach 


1. compute correlations of each protein with ADOS 
2. pick 15 proteins with the strongest correlation

**Need to explain following questions**
1. How do we utilize _ADOS_ score  
    ADOS scores: determined by psychological assessment and measure ASD severity.  
    6-23; measured for ASD group
    

```{r}
cor_test <- function(x, y){
  cor_out <- cor.test(x, y)
  tibble(estimate = cor_out$estimate,
         p.value = cor_out$p.value)
}

# use train_data
corr_result <- asd_training %>% select(-group) %>%
  pivot_longer(cols = -ados,
               names_to = 'protein',
               values_to = 'level') %>%
  group_by(protein) %>%
  summarize(correlation = cor_test(ados, level)) %>%
  unnest(correlation) %>%
  arrange(p.value)

corr_result_1 <- corr_result %>% head(15)
corr_result_1
```

```{r}
save(corr_result_1,corr_proteins, file = "corr_result.RData")
```


### Visual Assessment

```{r}
corr_proteins <- corr_result_1 %>% pull(protein)
corr_proteins
```

```{r}
a <- biomarker_clean %>% select(ados, all_of(corr_proteins))
```
```{r}
#pairs(ados ~ ., data = a)
```



## Random Forest 

We can control
1. number of trees T
2. number of predictors $m$ to choose at random for each tree
3. bootstrap sample size and method 
4. tree depth 

maybe doing hyperparameter later 


choose splits that optimize quantitative measures of _node homogeneity_

```{r}
library(randomForest)
# reproducibility
set.seed(101422)

# fit rf

asd_train_noados <- asd_training %>% select(-ados)

train_preds <- asd_train_noados %>% select(-group)
train_resp <- asd_train_noados %>% select(group)
train_resp$group <- as.factor(train_resp$group)


rf_out <- randomForest(x = train_preds, y = train_resp$group,
                       mtry = 100, ntree = 1000, 
                       importance = T)
```


```{r}
rf_proteins <- rf_out$importance %>% 
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  slice_max(MeanDecreaseGini, n = 15) %>%
  pull(protein)

rf_proteins
```

```{r}
save(rf_proteins, rf_out, file = "random_forst_result.RData")
```


## Choosing Core Panel 

```{r}
s1 <- intersect(t_test_proteins, corr_proteins)
s2 <- intersect(t_test_proteins, rf_proteins)
s3 <- intersect(rf_proteins, corr_proteins)

core_proteins <- s1 %>% union(s2) %>% union(s3)
core_proteins
```
Above is our core panel. 

```{r}
save(core_proteins, core_biomarker_clean, fit_logistic,
     confusion_table, accuracy_result_q3, file = "testing_result.RData")
```


## Constructing Logistic Regression 
Assumptions:
1. Observations are independent 
2. Probability of event is monotonic in each predictor
3. Mean-variance relationship following Bernoulli distribution

How to show above?


Maximum likelihood: find the parameter values for which the joint probability of the data is greatest according to the model.


```{r}
core_biomarker_clean <- asd_train_noados %>%
  select(group, all_of(core_proteins)) %>%
  mutate(is_asd = (group == 'ASD')) %>%
  select(-group) 

core_biomarker_clean
```

```{r}
# fit the logistic regression model
fit_logistic <- glm(is_asd ~ .,
                    family = 'binomial',
                    data = core_biomarker_clean)

fit_logistic %>%
  broom::tidy() %>%
  knitr::kable()

```


```{r}
confusion_table <- asd_testing %>% select(-ados) %>%
  modelr::add_predictions(fit_logistic, type = 'response') %>%
  mutate(pred_class = pred > 0.5) %>%
  select(group, pred_class) %>%
  mutate_all(~factor(.x, labels = c('TD', 'ASD'))) %>%
  table()

confusion_table
```

```{r}
library(yardstick)
# evaluate errors on test set
class_metrics <- metric_set(sensitivity, specificity, accuracy, roc_auc)

accuracy_result_q3 <- asd_testing %>% 
  modelr::add_predictions(fit_logistic, type = 'response') %>%
  class_metrics(estimate = factor(pred > 0.5),
                truth = factor(group == "ASD"),
                pred,
                event_level = 'second') #%>%
  #knitr::kable()


accuracy_result_q3

```

### ROC curve 


```{r}
a = asd_testing %>% select(-ados) %>%
  modelr::add_predictions(fit_logistic, type = 'response')

a%>% select(pred, group)
```


